---
title: "CRD and RBD"
author: "Ian Leboo"
date: "3/30/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,warning = TRUE)
```
#1. CRD 
##Example 1
The data contains 4 tropical food stuffs A, B, C, and D tried on 20 chicks. Analyze the data at 5% level of significance.
```{r}
library(readxl)
chicks <- read_excel("chics.xlsx")
```

Set the treatment column into factors
```{r}
chicks$Treatment<-as.factor(chics$Treatment)
```

Performing ANOVA
##Method 1
```{r}
mod1<-aov(Weight~Treatment, data = chics)
summary(mod1)
```

##Method 2
```{r}
chicksmodel<-lm(Weight~Treatment, data = chics)
anova(chicksmodel)
```
##LSD Test
```{r}
library(agricolae)
LSD.test(chicksmodel,"Treatment", console = TRUE)
```
##Example 2
The data contains the amount of electricity used in KWh in three towns. Test at 5% significance whether the amount of electricity used is the same in the 3 towns.
The steps are as example 1

```{r}
library(readxl)
electricity <- read_excel("electricity.xlsx")

#Set the treatment variable to factor
electricity$Town<-as.factor(electricity$Town)
model2<-aov(Electricity_Used~Town, data=electricity)
summary(model2)
```

#2. RBD

In this data, there are 4 blocks and 6 treatments.
we are required to test whether treatments and blocks differ significantly at 5% significance level
```{r}
library(readxl)
RBD <- read_excel("RBD.xlsx")
```

We will need to reshape the data into long format
```{r}
library(reshape2)
library(dplyr)
RBD_new<-RBD %>% 
  melt() %>% 
  mutate(Block=variable,
         Treatments=as.factor(Treatments)) %>% 
  select(-c(variable))
```

Performing ANOVA
```{r}
Model3<-aov(value~Treatments+Block, data=RBD_new)
summary(Model3)
```


